Verifying Spatial Properties of Array Computations

### 2017-04-17 oopsla17/algorithms.tex
edits to algorithms

\subsection{Checking code against specifications}
\label{subsec:checking}

In the model, we can regard both array indices and specifications as groups of
points in $N$-dimensional space. Notion of consistency is then intuitively
whether these two (potentially infinite) sets of points in space match or not.
This leads to a simple set based consistency function defined as
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
function to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost} modifier,
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices correspond exactly those defined by the
region in the specification, hence requiring set equality.

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite and set comparisons cannot be done by exhaustively
comparing the points of one with the other. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
it to \textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The
query belongs to to quantifier free linear arithmetic, which is a decidable
theory.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}. Performance of consistency checking
using satisfiability is irrelevant as the the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions conntected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
99\% of all array variables that we can assign a specification is X, and 99\% of
number of array terms that flows into such an assignment is Y. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

@@

>>>>

\subsection{Checking code against specifications}
\label{subsec:checking}
%
In the model, we can regard both array indices and specifications as sets of
points in $N$-dimensional space. The notion of consistency is then intuitively
whether these two (potentially infinite) sets of points match or not.
This leads to a simple set based consistency function defined as
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
function to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by the \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite and set comparisons cannot be done by exhaustively
comparing the points of one with the other. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
query is expressed in quantifier free linear arithmetic, which is decidable.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
99\% of all array variables that we can assign a specification is X, and 99\% of
number of array terms that flows into such an assignment is Y. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

@@

\begin{align*}
  \mathsf{coalesce1}(\mathcal{S}) & =
                                    \{\,S \cup T \mid S \in \mathcal{S}, 
                                    T \in \mathcal{S} \, \wedge \,
                                    \contig{S}{T} \} \\
  \mathsf{coalesce}(\mathcal{S}) & =
                                   \mathsf{coalesce}(\mathsf{coalesce1}(\mathcal{S}) \cup \mathcal{S})
                                   \; \cup \;
                                   \{ S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
\end{align*}
